Evolution of ORM Systems: Formal Foundations

### 2012-12-20 doc/description.tex
added texts

\maketitle
\section{ORM System}
The ORM system represents an application created using object oriented language and relational database. The components of such an application and theirs relations are in Figure \ref{fig:appStructure}. We define the application in our model as:

$$system = < Application, Database, \rho >$$
$instantiated : Class \times Application \rightarrow Boolean $ \\

The application and database has to be in consistent state according to a concrete ORM algorithm, otherwise the application cannot work properly. It means that during the evolution we have to respect a concrete ORM to produce consistent result too.

@@

All important components of the domain are shown in Figure \ref{fig:appStructure}. There are three static components (static because they do not define behavior, but state of the system): an application, a database schema and stored data. Then there are four dynamic components (they define transformations between components and transition between system states): an object-relational mapping is a transformation between application and database schema, then are two components for evolution of application and of database. (To simplify the model we will consider the database schema and data are evolved by one component). Last component is the most important one - it is component which maps the evolution from application level to relational database level. We call this component evolution object-relational mapping (eORM).

\section{Data Evolution in ORM System}
The Object-Relational system consist of an application and a database. The system has to be in consistent state, therefore the evolution of application has to be propagated correctly into database. Each evolution operation consist of two parts - applying evolution operation on application and on database. The mapping of the evolution change is called extended Object-Relational Mapping (eORM). The eORM mapping helps to assure consistency of the system by transforming application evolution into database schema and data. Each eORM function applied on an ORM system produces a new generation of the system. If the function cannot be applied on a system the system cannot be evolved by the function.

The state of a system is called generation and transformation from one state to another is called evolution. The state of the system defines a context for each evolution and it determines if the evolution is successful or not.

\begin{figure}
\begin{center}
	\includegraphics[scale=0.4]{./images/evolution_simple}
	\caption{The evolution of data changes the system on all levels}
\end{center}
	\label{fig:evolution}
\end{figure}

@@

There is a description of each component following. The description defines set of models of static components and dynamic components are defined as functions. There are several common parts of the model we can define in advance. The model uses a set $Labels$ which contains all possible identifiers for the model. Each model component which has a label ($label \in Labels$) is unambiguously identified in the model.

@@

\textbf{TODO: define consistency of rows and schema}

\section{Object-Relational Mapping}
There are defined functions for object-relational mapping used in the model. The purpose of a function is indicated in its name and should be obvious from its definition. These functions do not serve for full object-relational mapping, their purpose is to help with data evolution and its propagation from application to database.

@@

 \inference{ cardinality(p) > 1 \wedge ! mandatory(p) \\ \wedge \exists t : t = table(name(owningClass(p)), d)}{\begin{gathered}tables(d) \cup  table(name(p), primaryKey("Id"), column("value", \rho_t(type(p)), \varnothing), \\ foreignKey(name(p), name(owningClass(p)), NOTNULL) \\

@@

The new foreign key can cause inconsistency of the system when added to a table which contains data. This case is not handled by the mapping function above. We have to extend the function definition of a feature that allows the mapping between the data in both tables. The mapping is a function:
$$f_m : Property \rightarrow Instance \times Instance $$
%TODO handle instances

\section{Evolution}
\subsection{Evolution of Application}
\label{sec:appEvolution}

The evolution is defined as a set of operations which change structure of an application. The definition of an operation consists of two parts - a conditions of operation feasibility and operation impact on the application structure. We decide to use following transcription:

$$
operation = \inference{conditions$ $of$ $feasibility}{impact$ $on$ $application$ $structure}
$$
If the conditions of feasibility are not met, the operation cannot impact the application structure.
%TODO monade

@@

\subsection{ORM System Evolution}
Creational operations does not have impact on stored data they change the structure of the application and database by adding new objects into model.

\subsubsection{Creation of a System}
\paragraph{Add Class}

$$
\Phi_{addClass}(c, d) = \inference{ \forall t \in tables(d) : name(t) \neq name(\rho(c)) \\
	\forall a \in associations(c), \exists u \in  tables(d) : reference(a) = name(u) 
}{
\begin{gathered}
  \rho(c) \wedge \rho(singleProperties(c)) \wedge \rho(collectionProperties(c)) \\ \wedge \rho(associations(c))
\end{gathered}
}
$$
\paragraph{Add Property}
$$
\Phi_{addProperty}(p, d) = \begin{cases}
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge (! mandatory(p) \vee ! instantiated(owningClass(p))}{ \rho(p,d)} 
\\ \\ 
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge mandatory(p) \wedge instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered}
 \rho(p,d) \wedge \forall r \in data(d), reference(r) = t : pair(r) \\ \cup pair(name(p), defaultValue(p))
\end{gathered}
} 
\\ \\
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge mandatory(p) \wedge instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered}
 \rho(p,d) \wedge \forall r \in row(owningClass(p)) : data(d)\\ \cup row(name(p), pair("value", defaultValue(p)), \\ pair(name(p), value(primaryKey(r))) 
\end{gathered}
}

@@

\paragraph{Add Association}
$$
\Phi_{AddAssociation}(a, f_m, d) = \begin{cases}
\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = name(owningClass(a)) \\ \wedge name(t_t) = reference(a) \\ \wedge \not \exists f \in foreignKeys(t_t) : name(f) = name(a) \\ startCardinality(a) \leq 1}{\begin{gathered}
\rho(a,d) \wedge \forall r \in f_m(owningClass(a), reference(a)) : \\ data(t_t) \cup row(name(t_s), pair(name(a), \\ value(primaryKey(first(r)))) 
\end{gathered}
 }
 \\ \\
\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = name(owningClass(a)) \\ \wedge name(t_t) = reference(a) \\ \wedge \not \exists f \in foreignKeys(t_t) : name(f) = name(a) \\ startCardinality(a) > 1}{\begin{gathered}
\rho(a,d) \wedge \forall r \in f_m(owningClass(a), reference(a)),\\ t \in tables(d) : name(t) = name(a) : \\ data(t) \cup row(name(a), pair(name(t_s), \\ value(primaryKey(first(r)))) pair(name(t_t), \\ value(primaryKey(second(r)))))
\end{gathered}
 }

@@

\subsubsection{Deconstruction of a System}
\paragraph{Remove Class}
$$
\Phi_{removeClass}(c,d) = \inference{ \rho(c) \in tables(d) \\ \wedge \not \exists r \in data(d) : owningTable(r) = \rho(c) \\ \not \exists j \in tables(d), f \in foreignKeys(j) : reference(f) = \rho(c)  }{ tables(d) \setminus \rho(c)}
$$
%TODO remove by label

\paragraph{Remove Property}
$$
\Phi_{removeProperty}(p, d) = \begin{cases}
 \inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge \exists c \in columns(t) : name(c) = name(p) \\ \wedge \not \exists r \in data(d) : owningTable(r) = name(owningClass(p)) }{columns(t) \setminus c
} \\ \\
  \inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge \not \exists c \in columns(t) : name(c) = name(p) \\ \wedge \exists u \in tables(d) : name(u) = name(p) \\ \wedge \not \exists r \in data(d) : owningTable(r) = name(owningClass(p))}{ tables(d) \setminus u }
 \end{cases}
$$

\paragraph{Remove Association}
$$
\Phi_{removeAssociation}(a, d) =  \begin{cases}
 \inference{\exists t \in tables(d) : name(t) = name(owningClass(a))
 \\ \wedge \exists f \in foreignKeys(t) : name(f) = name(a)
 \\ \wedge \forall r \in data(d) : owningTable(r) \neq t}{ foreignKeys(t) \setminus f }
 \\ \\
 \inference{\exists t \in tables(d) : name(t) = name(a)
 \\ \wedge \exists f \in foreignKeys(t) : name(f) = name(owningClass(a))
 \\ \wedge \forall r \in data(d) : owningTable(r) \neq t}{ tables(d) \setminus t}
 \end{cases}
$$

>>>>

\maketitle
\begin{abstract}
The evolution of an application is a common issue during software development. The formal description of the evolution operations in context of  software created using an object oriented programming language and a relational database is introduced in this paper. Special attention is given to the data stored in databases (instances) and theirs consistency during evolution process.
\end{abstract}

\textbf{Keywords:} data evolution, data migration

\section{ORM System}
The ORM system represents an application created using object oriented language and relational database. The components of such an application and theirs relations are in Figure \ref{fig:appStructure}. We define the application in our model as:

$$system = < Application, Database, \rho >$$

The application and database has to be in consistent state according to a concrete ORM algorithm, otherwise the application cannot work properly. It means that during the evolution we have to respect a concrete ORM to produce consistent result.

@@

an object-relational mapping is a transformation between application and database schema

$instantiated : Class \times Application \rightarrow Boolean $ \\

\section{Data Evolution in ORM System}
The data evolution is a change of a state of the ORM system. The change of a state is described via evolution operations on application level, the change than has to be propagated into the rest of the system - database schema and data. The evolution process is in Figure \ref{fig:evolution}. There are three static components (static because they do not define behavior, but state of the system): an application, a database schema and stored data. Then there are two dynamic components (they define transition between system states): component for evolution of application and component for database evolution. To simplify the model we will consider the database schema and data are evolved by one component. Last component is the most important one - it is component which maps the evolution from application level to relational database level. We call this component evolution object-relational mapping (eORM).

%Each evolution operation consist of two parts - applying evolution operation on application and on database. The mapping of the evolution change is called extended Object-Relational Mapping (eORM). The state of a system is called generation and transformation from one state to another is called evolution. The state of the system defines a context for each evolution and it determines if the evolution is successful or not. The eORM mapping helps to assure consistency of the system by transforming application evolution into database schema and data. Each eORM function applied on an ORM system produces a new generation of the system. If the function cannot be applied on a system the system cannot be evolved by the function.

\begin{figure}
\begin{center}
	\includegraphics[scale=0.4]{./images/evolution_simple}
	\caption{The evolution of data changes the system on all levels. The figure shows all the components of the evolution process.}
\end{center}
	\label{fig:evolution}
\end{figure}

@@

A description of each component follows. Models of static components and dynamic components are defined, then ORM and eORM is defined. There are several common parts of the model we can define in advance. The model uses a set $Labels$ which contains all possible identifiers for the model. Each model component which has a label ($label \in Labels$) is unambiguously identified in the model.

@@

\textbf{TODO: define consistency of rows and schema}
\textbf{TODO: define operations such as instantiated}

\section{Object-Relational Mapping}
There are defined functions for object-relational mapping used in the model in this section. The purpose of a function is indicated in its name and should be obvious from its definition. These functions do not serve for full object-relational mapping, their purpose is to help with data evolution (see Section \ref{sec:eorm}) and its propagation from application to database.

@@

 \inference{ cardinality(p) > 1 \wedge ! mandatory(p) \\ \wedge \exists t : t = table(name(owningClass(p)), d)}{\begin{gathered}tables(d) \cup  table(name(p), primaryKey("Id"), \\ column("value", \rho_t(type(p)), \varnothing), \\ foreignKey(name(p), name(owningClass(p)), NOTNULL) \\

@@

\section{Evolution}
\label{sec:eorm}
There are introduced operations for evolution of an application first as they are a source for the evolution of the whole ORM system. The mapping of the application evolution into a database schema and stored data are introduced later in this section too.

\subsection{Evolution of Application}
\label{sec:appEvolution}

The evolution of an application is defined as a set of operations which change a structure of the application. The definition of an operation consists of two parts - a conditions of operation feasibility and operation impact on the application structure. We decide to use following transcription:

$$
operation = \inference{conditions$ $of$ $feasibility}{impact$ $on$ $application$ $structure}
$$
If the conditions of feasibility are not met, the operation cannot impact the application structure.

@@

\subsection{Extended Object-Relation Mapping}
The extended object-relational mapping maps the evolution of the application level into the database schema and stored data. This section introduces the mapping for evolution operations introduced earlier.

\subsubsection{Creation of a System}
\paragraph{Add Class} The $AddClass$ operation changes the structure of the database schema only.
$$
\Phi_{AC}(c, d) = \inference{ \forall t \in tables(d) : name(t) \neq name(\rho(c)) \\
	\forall a \in associations(c), \exists u \in  tables(d) : reference(a) = name(u) 
}{
\begin{gathered}
  \rho(c) \wedge \rho(singleProperties(c)) \wedge \rho(collectionProperties(c)) \\ \wedge \rho(associations(c))
\end{gathered}
}
$$

\paragraph{Add Property} The function $AddProperty$ on level of database does not change the structure only, but it has an impact on stored data too. This is because a mandatory property can be added into a class with already existing instances, then a default value has to be added to all existing instances.
$$
\Phi_{AP}(p, d) = \begin{cases}
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge (! mandatory(p) \vee ! instantiated(owningClass(p))}{ \rho(p,d)} 
\\ \\ 
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge mandatory(p) \wedge instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered}
 \rho(p,d) \wedge \forall r \in data(d), reference(r) = t : pair(r) \\ \cup pair(name(p), defaultValue(p))
\end{gathered}
} 
\\ \\
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge mandatory(p) \wedge instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered}
 \rho(p,d) \wedge \forall r \in row(owningClass(p)) : data(d)\\ \cup row(name(p), pair("value", defaultValue(p)), \\ pair(name(p), value(primaryKey(r))) 
\end{gathered}
}

@@

\paragraph{Add Association} Adding an association into a database is complicated because of already existing instances. When an association between classes is added the instances of both classes has to be paired. This pairing is provided by the function $f_m$:
$$f_m : Property \rightarrow Instance \times Instance $$

\textbf{TODO: define the features of the $f_m$}

The mapping of the $AddAssociation$ function is defined as:
$$
\Phi_{AA}(a, f_m, d) = \begin{cases}
\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = name(owningClass(a)) \\ \wedge name(t_t) = reference(a) \\ \wedge \not \exists f \in foreignKeys(t_t) : name(f) = name(a) \\ startCardinality(a) \leq 1}{\begin{gathered}
\rho(a,d) \wedge \forall r \in f_m(owningClass(a), reference(a)) : \\ data(t_t) \cup row(name(t_s), pair(name(a), \\ value(primaryKey(first(r)))) 
\end{gathered}
 }
 \\ \\
\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = name(owningClass(a)) \\ \wedge name(t_t) = reference(a) \\ \wedge \not \exists f \in foreignKeys(t_t) : name(f) = name(a) \\ startCardinality(a) > 1}{\begin{gathered}
\rho(a,d) \wedge \forall r \in f_m(owningClass(a), reference(a)),\\ t \in tables(d) : name(t) = name(a) : \\ data(t) \cup row(name(a), pair(name(t_s), \\ value(primaryKey(first(r)))) pair(name(t_t), \\ value(primaryKey(second(r)))))
\end{gathered}
 }

@@

\subsubsection{Deconstruction of a System}
\paragraph{Remove Class}
$$
\Phi_{RC}(c,d) = \inference{ \rho(c) \in tables(d) \\ \wedge \not \exists r \in data(d) : owningTable(r) = \rho(c) \\ \not \exists j \in tables(d), f \in foreignKeys(j) : reference(f) = \rho(c)  }{ tables(d) \setminus \rho(c)}
$$
%TODO remove by label

\paragraph{Remove Property}
$$
\Phi_{RP}(p, d) = \begin{cases}
 \inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge \exists c \in columns(t) : name(c) = name(p) \\ \wedge \not \exists r \in data(d) : owningTable(r) = name(owningClass(p)) }{columns(t) \setminus c
} \\ \\
  \inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge \not \exists c \in columns(t) : name(c) = name(p) \\ \wedge \exists u \in tables(d) : name(u) = name(p) \\ \wedge \not \exists r \in data(d) : owningTable(r) = name(owningClass(p))}{ tables(d) \setminus u }
 \end{cases}
$$

\paragraph{Remove Association}
$$
\Phi_{RA}(a, d) =  \begin{cases}
 \inference{\exists t \in tables(d) : name(t) = name(owningClass(a))
 \\ \wedge \exists f \in foreignKeys(t) : name(f) = name(a)
 \\ \wedge \forall r \in data(d) : owningTable(r) \neq t}{ foreignKeys(t) \setminus f }
 \\ \\
 \inference{\exists t \in tables(d) : name(t) = name(a)
 \\ \wedge \exists f \in foreignKeys(t) : name(f) = name(owningClass(a))
 \\ \wedge \forall r \in data(d) : owningTable(r) \neq t}{ tables(d) \setminus t}
 \end{cases}
$$
